Using Optimal Time Step Selection to Boost the Accuracy of Finite Difference Methods for a Class of Time Dependent Partial Differential Equations

### 2008-05-10 papers/OTSPDE/optimal_time_steps_for_pdes.tex
OTSPDE:  first completed draft of paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First draft: 2008/02/??, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\author{
Kevin T. Chu\footnotemark[2]
}

@@

@@

\begin{abstract}
In this article, we discuss a novel technique based on optimal time step 
selection for boosting low-order finite difference schemes
into higher-order numerical methods for time dependent partial differential
equations (PDEs).  We demonstrate the utility of this technique on several 
classical hyperbolic and parabolic problems in one and more space dimensions 
and explain the observed orders of convergence through straightforward numerical
analysis arguments.
** HIGH-ORDER PDES AND IRREGULAR DOMAINS ***
\end{abstract}

\begin{keywords}
optimal time step, finite difference methods, high-order accuracy, 
time dependent PDEs, initial-value problems
\end{keywords}

@@

\section*{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical schemes~\cite{gibour_2005,??}.  
Typically, high-order accuracy is achieved by constructing
schemes that have high formal orders of accuracy.  However, high-order 
accurate numerical solutions can also be obtained by using formally low-order 
schemes in clever ways.  When possible, the latter approach can be a powerful 
way to boost the accuracy of a numerical method \emph{without} introducing too 
much additional algorithmic (and programming) complexity.

@@

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This standard approach for developing numerical methods for PDEs 
has its roots in the classical numerical analysis of PDEs, which tends to 
individually analyze the discretization of the terms in the PDE.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for \emph{all} 
independent variables.  Unfortunately, this common approach for developing 
and analyzing numerical methods is often suboptimal because it neglects the 
potential for achieving high order of accuracy using only simple finite 
difference schemes based on compact, low-order stencils.

Perhaps the greatest weakness of the standard method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides all 
of the details about higher-order errors.  The information lost in the ``big-O" 
notation eliminates the possibility of reducing numerical error by taking 
advantage of relationships between the terms in the error.  The modified 
nine-point formula for the Poisson equation (also known under the name 
Mehrstellen)~\cite{iserles_book} and the Lax-Wendroff scheme for the 
advection equation~\cite{leveque_book_1992, leveque_book_2002, gko_book} 
demonstrate the power of explicitly retaining higher-order terms in the 
discretization error when analyzing numerical schemes.  For both of these 
methods, examination of higher-order terms allows low-order methods to be 
transformed into high-order methods with only minor modifications to the 
numerical scheme.

@@

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection in detail and 
presenting the analysis that shows how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
to improve the accuracy of simple finite difference schemes for the linear 
advection equation and diffusion equation in multiple space dimensions.  
To illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both 
the regular and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we provide some concluding remarks on 
the applicability and limitations of optimal time step selection for the 
numerical solution of general time dependent PDEs.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section we present and analyze the 
technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
$\frac{\partial u}{\partial t} = \ldots$, where the right hand side only 
involves functions of the independent variable and its spatial derivatives.  
Because this class of time dependent PDEs is so prevalent and important in 
many problems of applied mathematics, restricting our attention to problems 
of this form is not a serious limitation.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation to a 
time dependent PDE of the form
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $u$ is the solution to the PDE, $n$ is the spatial order of the 
PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the FD 
approximation by optimizing the time step, there is no reason to explicitly
construct the scheme to be high-order.  In fact, using a high-order FD scheme 
only complicates the analysis required to calculate the optimal time 
step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the FD approximation, explicitly keeping track 
of the leading-order terms.  In contrast to standard error analysis, we do 
not sweep all of the errors under ``big-O'' notation.  Rather, we take 
advantage of the fact that derivation of the leading-order terms in the 
error is straightforward for FD schemes through the use of Taylor series 
expansions.  Because each term in the error is proportional to a partial 
derivative of the solution, we can attempt to use the PDE to relate the 
leading-order terms to each other.  If we are fortunate (which in many cases 
we are), this careful analysis yields a direct relationship between the 
leading-order spatial and temporal errors that allows them to be combined as 
a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $P$ is a polynomial in the numerical parameters, $L$ is a differential 
operator, $u$ is the solution to the PDE, and we have assumed that the 
grid spacing is equal in all spatial directions\footnote{In general, the grid 
spacing need not be equal in all spatial directions.  For some problems, it is 
only possible to calculate an optimal time step if the grid spacing is
unequal between the different spatial directions.}.  By selecting the 
numerical parameters so that $P = 0$, we can completely eliminate the 
leading-order term in the discretization error.  Because time dependent PDEs 
are often solved by stepping in time, it is natural to choose the time step 
to be a function of the grid spacing.  The time step selected in this way 
is the \emph{optimal time step} for the finite difference scheme.

An important class of FD schemes for time dependent PDEs are those based
on the method of lines.   When the method of lines approach is used for 
(\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
errors, and $\alpha$ and $\beta$ are positive constants that depend on the 
details of the PDE and FD scheme.  Setting $P = 0$, we find that the optimal 
time step is given by\footnote{The $\dt = 0$ root of $P$ is discarded 
because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Note that the optimal time step 
has the same functional dependence on the grid spacing as a time step 
that is satisfies a stability constraint.

The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the FD approximation of the spatial derivative 
operator acting on ${\mathbf u}$.  The second term in the local error arises 
directly from temporal discretization errors associated with the numerical 
scheme used to integrate the system of ODEs in time.  The first term arises 
because $F$ is only an approximation of the spatial derivative operators in 
the PDE.  As a result, the system of ODEs~(\ref{eq:method_of_lines}) has an 
error due to the spatial discretization at \emph{every} point in time.  For 
a spatial discretization error of order $r$, the error in 
(\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the error 
accumulated during each time step due to spatial discretization error is 
$O(\dt \dx^r)$ regardless of how the system of ODEs is solved 
numerically.  Because the method of lines is so frequently used to construct FD 
approximations for time dependent PDEs, we will focus solely on FD schemes 
of this type for the remainder of the paper in order to keep the discussion 
simple.  It is straightforward to extend the results we obtain to more general 
FD schemes.

So far in our discussion, we have completely ignored the question of whether 
the FD scheme is stable when the time step is set to $\dto$.  When
the optimal time step is greater than the smallest stable time step, the 
method breaks down.  Fortunately, when stability issues arise for explicit 
time integration schemes, we can usually switch to an implicit (or 
semi-implicit) time integration scheme with the same order of accuracy for 
which the optimal time step is stable.  Note that the depending on how
the original FD scheme is modified, a new optimal time step may need to be 
calculated.

\subsubsection*{PDEs in Higher Spatial Dimensions 
                \label{sec:ots_higher_spatial_dims}}
It is important to emphasize that optimal time selection is applicable to 
PDEs in any number of spatial dimensions.  While problems in higher dimensions 
generally require a more careful choice of the finite difference stencil used 
to approximate spatial derivatives in order to be able to combine 
leading-order terms in the discretization error, there are no fundamental 
barriers that prevent the use of optimal time step selection.

Domain geometry of the domain, which primarily impacts the way boundary 
conditions are imposed, is an important feature of problems in more than one 
spatial dimension.  While it is usually straightforward to handle problems on 
regular domains, problems on irregular domains can pose a greater challenge 
because the location of the domain boundary may not be explicitly specified
and grid points near the boundary require special consideration.  Fortunately, 
because optimal time step selection is based on simple, localized finite 
difference stencils, irregular domains do not create any serious difficulties.  
Assuming that a `ghost cell' or `ghost value' 
approach~\cite{fedkiw_1999, osher_fedkiw_book, gibou_2005} 
is used to impose boundary conditions, the construction and analysis of finite 
difference schemes and calculation of optimal time steps is identical for 
problems on both regular and irregular domains.  For problems on 
irregular domains, the key issue is how values at ghost cell points should 
be calculated.  Once this issue has been addressed, devising high-order 
accurate methods for PDEs on irregular domains is simply a matter of applying 
optimal time step selection to the finite difference scheme used for problems 
on regular domains.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
To analyze the accuracy of the FD scheme for~(\ref{eq:model_PDE}) when the 
optimal time step is used (for a given grid spacing), we examine the remaining 
terms of the discretization error in the standard way (\ie ``big-O'' notation).
Elimination of the leading term in the discretization error leaves a local 
error of the form $O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ 
and $(q+1) > (s+1)$ are the orders of the spatial and temporal discretization 
errors, respectively, after the leading-order terms in the error have been 
removed.  Like the orders of the leading-order errors in 
(\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
error is a direct consequence of the method of lines approach to discretizing 
the PDE.

@@

When interpreting~(\ref{eq:global_error_ots}), it is important to recognize
that $\dx$ and $\dt$ are \emph{not} independent because the time 
step has been set to $\dto$; there is really only one parameter, 
$\dx$, that controls the numerical accuracy of the FD scheme.  This is an 
important point that we will address further in the next section.

@@

Optimal time step selection fits into this general framework for developing
numerical methods for time dependent PDEs.  Because the optimal time step 
is calculated explicitly as a function of the grid spacing, the technique
of optimal time step selection places an even more restrictive constraint
on $\dt$ than stability considerations.  Thus, like numerical
methods with stability constraints, the accuracy of an FD scheme using 
the optimal time step is completely controlled by the grid spacing.  The 
order of accuracy of a FD scheme advanced in time using $\dto$ 
depends on the functional dependence the optimal time step on $\dx$.
For the method of lines, we know that $\dto = O(\dx^{r/s})$, 
so the global error derived in the previous section reduces to 
\beq
O \left( \dx^{\min(p,rq/s)} \right).
\label{eq:global_error_ots_simplified}
\eeq

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
In general, $r/s > 1$ and yields an optimal time step that is on the order
of the time step allowed by the stability constraint for an explicit method.
As a result, using the optimal time step to advance the FD scheme in time
may seem to be overly restrictive and lead to poor computational performance.
However, these apparent drawbacks are more than compensated by the boost in 
the order of accuracy.  For example, when solving the 1D diffusion equation 
using a forward Euler time integration scheme with a second-order central 
difference stencil for the Laplacian, we obtain a fourth-order scheme 
when $\dt$ is set equal to the optimal time step.  As
Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order FD stencils.  The performance gain for problems in higher 
spatial dimensions is even more impressive.  As we can see in 
Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tbh]
\caption{\label{tab:comp_perf_vs_err} 
   Computational cost as a function of the global numerical error $E$
   for various FD schemes that numerically solve the 1D diffusion equation.  
   For all FD schemes, the standard second-order central difference 
   stencil is used to discretize the Laplacian, and the numerical 
   solution is assumed to be computed over a fixed time interval.  
A less negative power of $E$ 
   for the memory and compute time means that a FD scheme is more 
   computationally efficient.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The 
   time step for the forward/backward Euler and Crank-Nicholson schemes 
   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
   time step for the backward Euler scheme is chosen so that the temporal 
   error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the implicit methods.}  \\
  \hline 
  Forward Euler    & $O\left( E^{1/2} \right)$ 
                   & $O\left (E \right)$ 
                   & $O\left( E^{-1/2} \right)$ 
                   & $O\left( E^{-3/2} \right)$ \\
  Backward Euler   & $O\left( E^{1/2} \right)$ 
                   & $O\left( E \right)$ 
                   & $O\left( E^{-1/2} \right)$ 
                   & $O\left( E^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( E^{1/2} \right)$ 
                   & $O\left( E^{1/2} \right)$ 
                   & $O\left( E^{-1/2} \right)$ 
                   & $O\left( E^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( E^{1/4} \right)$ 
                   & $O\left( E^{1/2} \right)$ 
                   & $O\left( E^{-1/4} \right)$ 
                   & $O\left( E^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First draft: 2008/05/??, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\author{
Kevin T. Chu\footnotemark[2] \footnotemark[3]
}

@@

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[3]{Institute of High Performance Computing, A*STAR, Singapore, Singapore}

@@

\begin{abstract}
In this article, we present a novel technique for obtaining high-order
accurate numerical solutions to time dependent partial differential 
equations using only formally low-order finite difference stencils and
time integration schemes.  By using an optimally chosen the time step for a 
finite difference scheme (and possibly adding a few correction terms), we 
are able to boost the order of accuracy of low-order finite difference 
schemes.  Through simple numerical analysis arguments, we explain the origin 
of the boost in accuracy and calculate the computational cost of the
high-order methods obtained via optimal time step (OTS) selection.  We 
demonstrate the utility of OTS selection on a wide array of classical PDEs
in one and more space dimensions.   We find that OTS selection is applicable 
to linear and semilinear PDEs and that it is easily extended to problems 
on irregular domains. 
\end{abstract}

\begin{keywords}
optimal time step, finite difference methods, high-order accurate numerical 
methods, time dependent PDEs
\end{keywords}

@@

\section*{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical schemes~\cite{bruger_2005, gibou_2005, ito_2005, shukla_2005, 
shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that have high formal orders of accuracy.  However, high-order 
accurate numerical solutions can also be obtained by using formally low-order 
schemes in clever ways.  When possible, the latter approach can be a powerful 
way to boost the accuracy of a numerical method \emph{without} introducing too 
much additional algorithmic (and programming) complexity.

@@

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This conventional approach for developing numerical methods for 
PDEs has its roots in the classical numerical analysis of PDEs, which tends to 
individually analyze the discretization of the terms in the PDE.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for \emph{all} 
independent variables.  Unfortunately, this common approach for developing 
and analyzing numerical methods is often suboptimal because it neglects the 
potential for achieving high order of accuracy using only simple finite 
difference schemes based on compact, low-order stencils.

Perhaps the greatest weakness of the conventional method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides all 
of the details about higher-order errors.  The information lost in the ``big-O" 
notation eliminates the possibility of reducing numerical error by taking 
advantage of relationships between the terms in the error.  The modified 
nine-point formula for the Poisson equation (also known under the name 
Mehrstellen)~\cite{iserles_book} and the Lax-Wendroff scheme for the 
advection equation~\cite{leveque_book_1992, leveque_book_2002, gko_book} 
demonstrate the power of explicitly retaining higher-order terms in the 
discretization error when analyzing numerical schemes.  For both of these 
methods, examination of higher-order terms allows low-order methods to be 
transformed into high-order methods with only minor modifications to the 
numerical scheme.

@@

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection in detail and 
presenting the analysis that shows how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
to improve the accuracy of simple finite difference schemes for the linear 
advection equation and diffusion equation in multiple space dimensions.  
To illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both 
the regular and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section we present and analyze the 
technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important in 
many problems of applied mathematics, restricting our attention to problems 
of this form is not a serious limitation.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $u$ is the solution to the PDE, $n$ is the spatial order of the 
PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the FD 
approximation by optimizing the time step, there is no reason to explicitly
construct the scheme to be high-order.  In fact, using a high-order FD scheme 
only complicates the analysis required to calculate the optimal time 
step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the FD approximation, explicitly keeping track 
of the leading-order terms.  In contrast to conventional error analysis, we 
do not sweep all of the errors under ``big-O'' notation.  Rather, we take 
advantage of the fact that derivation of the leading-order terms in the 
error is straightforward for FD schemes through the use of Taylor series 
expansions.  Because each term in the error is proportional to a partial 
derivative of the solution, we can attempt to use the PDE to relate the 
leading-order terms to each other.  If we are fortunate (which we are
for many finite difference schemes, this careful analysis yields a direct 
relationship between the leading-order spatial and temporal errors that allows 
them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a polynomial in the numerical 
parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
$P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with the maximal order 
of accuracy for a given FD scheme.

An important class of FD schemes for time dependent PDEs are those based
on the method of lines.   When the method of lines approach is used for 
(\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
errors, and $\alpha$ and $\beta$ are positive constants that depend on the 
details of the PDE and FD scheme.  Setting $P = 0$, we find that the optimal 
time step is given by\footnote{The $\dt = 0$ root of $P$ is discarded 
because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
has a similar functional dependence on the grid spacing as a time step that 
satisfies a stability constraint.

Thus far, we have completely ignored the question of whether the FD scheme is 
stable when the time step is set to $\dto$.  Not surprisingly, OTS selection 
loses its value, when the optimal time step is greater than the smallest 
stable time step for an explicit scheme.  Fortunately, we can usually switch 
to an implicit (or semi-implicit) time integration scheme with the same order 
of accuracy for which the optimal time step is stable.  Note that the 
depending on how the original FD scheme is modified, a new optimal time step 
may need to be calculated.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the FD approximation of the spatial derivative 
operator acting on ${\mathbf u}$.  The second term in the local error arises 
directly from temporal discretization errors associated with the numerical 
scheme used to integrate the system of ODEs in time.  The first term arises 
because $F$ is only an approximation of the spatial derivative operators in 
the PDE.  As a result, the system of ODEs~(\ref{eq:method_of_lines}) has an 
error due to the spatial discretization at \emph{every} point in time.  For 
a spatial discretization error of order $r$, the error in 
(\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the error 
accumulated during each time step due to spatial discretization error is 
$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
is solved numerically.  Because the method of lines is so frequently used to 
construct FD approximations for time dependent PDEs, we will focus solely on 
FD schemes of this type for the remainder of the paper to keep the discussion 
simple.  It is straightforward to extend the results we obtain to more general 
FD schemes.

To analyze the accuracy of the FD scheme for~(\ref{eq:model_PDE}) when the 
optimal time step is used (for a given grid spacing), we examine the 
higher-order terms in the discretization error following the conventional 
approach (\ie ``big-O'' notation).  Elimination of the leading term in the 
discretization error leaves a local error of the form 
$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
of the spatial and temporal discretization errors, respectively, after the 
leading-order terms in the error have been removed.  As for the leading-order 
error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
error is a direct consequence of the method of lines approach to discretizing 
the PDE.

@@

It is important to emphasize that $\dx$ and $\dt$ are \emph{not} independent 
in~(\ref{eq:global_error_ots}) because the time step must be set to $\dto$ 
in order to derive the error estimate.  In other words, there is really only 
one parameter, $\dx$, that controls the numerical accuracy of the FD scheme.  
This important point will be discuss in further detail in the next section.

@@

Optimal time step selection fits into this general framework for controlling
the error of numerical methods for time dependent PDEs.  Because the optimal 
time step is calculated explicitly as a function of the grid spacing, the 
technique of optimal time step selection places an even more restrictive 
constraint on $\dt$ than stability considerations.  Thus, like numerical
methods with stability constraints, the accuracy of an FD scheme using 
the optimal time step is completely controlled by the grid spacing.  The 
order of accuracy of a FD scheme advanced in time using $\dto$ 
depends on the functional dependence the optimal time step on $\dx$.
For the method of lines, we know that $\dto = O(\dx^{r/s})$, 
so the global error derived in the previous section reduces to 
\beq
O \left( \dx^{\min(p,rq/s)} \right).
\label{eq:global_error_ots_simplified}
\eeq

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
For many problems, $r/s > 1$ and yields an optimal time step that is on the 
order of the time step allowed by the stability constraint for an explicit 
method.  As a result, using the optimal time step to advance the FD scheme in 
time may seem to be overly restrictive and lead to poor computational 
performance.  However, these apparent drawbacks are more than compensated by 
the boost in the order of accuracy.  For example, when solving the 1D 
diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, we obtain a 
fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order FD stencils.  The performance gain for problems in higher 
spatial dimensions is even more impressive.  As we can see in 
Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tbh]
\caption{\label{tab:comp_perf_vs_err} 
   Computational cost of various FD schemes for numerically solving the 1D 
   diffusion equation as a function of the global numerical error $e$.
   For all FD schemes, the standard second-order central difference 
   stencil is used to discretize the Laplacian, and the numerical 
   solution is assumed to be computed over a fixed time interval.  
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The 
   time step for the forward/backward Euler and Crank-Nicholson schemes 
   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
   time step for the backward Euler scheme is chosen so that the temporal 
   error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the implicit methods.}  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\begin{table}[tbh]
\caption{\label{tab:comp_perf_vs_dim}
   Computational cost of the forward Euler scheme with optimal time step and 
   the Crank-Nicholson method for solving the diffusion equation as a function 
   of the global numerical error $e$.
   For both FD schemes, second-order, isotropic stencils are used to discretize 
   the Laplacian.  For a problem in $d$ spatial dimensions, the memory 
   requirement for each scheme is estimated to be $O \left(1/\dx^d \right)$, 
   and the computation time is approximated as 
   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, which assumes that we 
   are using an efficient $O(N) = O \left( 1/\dx^d \right)$ algorithm to 
   invert the matrices that arise for the Crank-Nicholson method.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|c|c|}
  \hline
  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( e^{-1/4} \right)$ 
      & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ \\ 
  $2$ & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-3/2} \right)$ \\ 
  $3$ & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-5/4} \right)$ 
      & $O\left( e^{-3/2} \right)$ 
      & $O\left( e^{-2} \right)$ \\
  $d$ & $O\left( e^{-d/4} \right)$ 
      & $O\left( e^{-(d+2)/4} \right)$ 
      & $O\left( e^{-d/2} \right)$ 
      & $O\left( e^{-(d+1)/2} \right)$ \\ 
  \hline 
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\subsection{\label{sec:ots_general_1d_pdes} 
            OTS Selection for General PDEs in One Space Dimension} 
In the previous section, we analyzed the technique of optimal time step 
selection in the context of a simple model PDE in one space dimension.  
For more general PDEs, the approach of carefully choosing the time step to 
eliminate leading order terms in the error is still useful, but it needs to 
be augmented to achieve a boost in the order of accuracy.  In addition to 
using the optimal time step, correction terms may need to be added to the 
finite difference scheme.  In this section, we discuss these correction terms 
in the context of general constant coefficient, linear PDEs and semilinear
PDEs.
